---
title: "Fracking Oil Loss"
author: "Hannah Marszalek"
date: "2024-06-03"
output: html_document
---

# Oil Companies in Texas Lose Thousands of Barrels Per Year

### The mining, quarrying, and oil and gas extraction industry is one of Texas' most profitable, yet every year many companies experience major losses.

Hannah Marszalek

June 5, 2024

Every year, oil companies in Texas lose thousands of barrels of liquid hydrocarbon (crude oil and gas well liquid) while fracking, amounting to a recorded total of 24,489,864 gallons lost since 2009 --- enough oil to fill 37 Olympic swimming pools. The companies are required to [fill out a form known as an H-8](https://www.rrc.texas.gov/media/fscnvxwt/h-8f.pdf) and submit it to the Railroad Commission of Texas each time such an event occurs, but this information leaves room for plenty of questions regarding the effects and frequency of such events.

The number of barrels lost has been on a steady decline since 2017 with 6,130 barrels lost in 2023, but this does not make these losses any less significant considering only 3,729 (about 60%) of these barrels were recovered. Of the 10,585 cases where equipment failure or corrosion was cited as the cause of the incident, only 522 mentioned mentioned equipment repair as their remedial measures taken. This points to widespread issues with the existing fracking equipment, which creates even more opportunity for such losses to occur.

The companies that have lost the most since 2009 are NFR Energy, Amid Silver Dollar Pipeline, Endeavor Energy Resources, Epic Consolidated Operations,and Anadarko E&P Onshore. A particularly interesting aspect about these being the biggest losers is that three of these five companies have very incomplete loss records. While a company is only required to fill out an H-8 form if they have a loss, it seems strange that those who had such large amounts of losses between only one or two years did not have any losses from any other years.

If I were to continue forward with only focusing on Texas, the target audience would be the people who live there so they can become more aware of what is happening in their community since there may be long-lasting effects from these losses. If I were to broaden the focus to more states, the target audience would be anyone who lives near those fracking sites. For interviews, I could try to speak to someone from one of the companies that are losing the most, an expert on the environmental effects of these oil spills, and someone who lives near these fracking locations to see how they feel about it.

I would like to gather information that will help me gain a better understanding of the remedial process following a spill and information that sheds light on the incidents that have occurred in the counties with the highest numbers of losses. For example, 26,199 barrels of crude oil were lost in Crockett County in 2019, and since this was the highest loss in that county out of all the years, it makes me wonder what happened that year that made it so much worse. Regarding the remedial process, many of the entries say there was "soil remediation" but they do not provide any detail as to what that actually entails. Furthermore, information on how much oil was produced these years would provide context to the significance of these losses.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(readxl)
library(janitor)
library(tidyverse)
library(kableExtra)

oil_loss <- read_excel("combined_h8.xlsx") %>% 
  clean_names()
```

## Section 1: Net Loss and Amount Recovered Per Year

### Net Loss

```{r}
#net loss of crude oil by year
loss_per_year <- oil_loss %>% 
  group_by(master_year) %>% 
  summarize(total_loss = sum(net_loss, na.rm=TRUE))

#amount recovered
sum(oil_loss$recovered) #the number of recovered barrels is more than the net loss of barrels, but still not great

#recovered crude oil by year
recovered_per_year <- oil_loss %>% 
  group_by(master_year) %>% 
  summarize(total_recovered = sum(recovered, na.rm=TRUE))

#while net loss has been on the decline in recent years, in 2023, still only about 60% of lost oil was recovered (3453/5519), as can be seen here:
oil_loss %>% 
  filter(master_year==2023) %>% 
  summarize(total_lost = sum(net_loss, na.rm=TRUE))

oil_loss %>% 
  filter(master_year==2023) %>% 
  summarize(total_recovered = sum(recovered, na.rm=TRUE))
```

### Amount Recovered

```{r}
#creating a new column to calculate what percent of the losses typically gets recovered
pct_recovered <- oil_loss %>% 
  select(master_year, operator_number, county, type_liquid, gross_loss, recovered, net_loss) %>% 
  mutate(pct_recovered = (recovered/gross_loss)*100)

#hypothesis: if the number of barrels is fewer, they are less likely to recover as much. (doesn't seem to actually be the case as a blanket statement)
pct_recovered %>% 
  filter(recovered==0) %>% 
  summarize(avg_loss = mean(gross_loss, na.rm=TRUE))

pct_recovered %>%
  filter(pct_recovered==100) %>%
  summarize(avg_loss = mean(gross_loss, na.rm=TRUE))
```

```{r}
#net loss graph
ggplot(loss_per_year, aes(x=master_year, y=total_loss)) +
  geom_line() +
  labs(title = "Net loss of crude oil at an all time low since 2009")
```

```{r}
#recovered barrels graph
ggplot(recovered_per_year, aes(x=master_year, y=total_recovered)) +
  geom_line() +
  labs(title = "Number of recovered barrels of crude oil on the decline")
```

## Section 2: A Closer Look at the Companies

### Calculating the number of individual reported incidents per company

```{r}
#totals across all years
companies <- oil_loss %>% 
  group_by(operator_number) %>% 
  count() %>% 
  arrange(desc(n))

head(companies)

#totals by year and company
companies_years <- oil_loss %>% 
  group_by(operator_number, master_year) %>% 
  count() %>% 
  arrange(desc(n))

head(companies_years)
```

### Finding the net loss of barrels for all incidents per company

```{r}
sum(oil_loss$net_loss) * 42
#24,489,864 gallons of crude oil and gas well liquids have been lost since 2009!

#net loss per company
net_loss <- oil_loss %>% 
  select(operator_number, net_loss) %>% 
  group_by(operator_number) %>%
  summarize(total_barrels = sum(net_loss, na.rm=TRUE)) %>% 
  arrange(desc(total_barrels))

#looking at average lost per company
oil_loss %>%
  select(operator_number, net_loss) %>% 
  group_by(operator_number) %>%
  summarize(avg_barrels = mean(net_loss, na.rm=TRUE)) %>% 
  arrange(desc(avg_barrels))
```

### Texas' top producing companies

<https://www.rrc.texas.gov/oil-and-gas/research-and-statistics/operator-information/texas-oil-and-gas-producers-by-rank-2023/>

<iframe src="https://flo.uri.sh/visualisation/18240706/embed" title="Interactive or visual content" class="flourish-embed-iframe" frameborder="0" scrolling="no" style="width:100%;height:600px;" sandbox="allow-same-origin allow-forms allow-scripts allow-downloads allow-popups allow-popups-to-escape-sandbox allow-top-navigation-by-user-activation">

</iframe>

::: {style="width:100%!;margin-top:4px!important;text-align:right!important;"}
<a class='flourish-credit' href='https://public.flourish.studio/visualisation/18240706/?utm_source=embed&utm_campaign=visualisation/18240706' target='_top' style='text-decoration:none!important'><img src="https://public.flourish.studio/resources/made_with_flourish.svg" alt="Made with Flourish" style="width:105px!important;height:16px!important;border:none!important;margin:0!important;"/> </a>
:::

```{r}
#top 5 by oil volume rank: Pioneer Natural Resources, Diamondback, Endeavor Energy Resources, EOG Resources, XTO Energy
top_5_producers <- oil_loss %>% 
  filter(operator_number=="PIONEER NATURAL RES. USA, INC."|operator_number=="DIAMONDBACK E&P LLC"|operator_number=="ENDEAVOR ENERGY RESOURCES L.P."|operator_number=="EOG RESOURCES, INC."|operator_number=="XTO ENERGY INC.") %>%
  group_by(operator_number, master_year) %>% 
  summarize(total_lost = sum(net_loss, na.rm=TRUE))

top_5_producers <- top_5_producers %>% 
  pivot_wider(names_from=operator_number, values_from=total_lost)
 
top_5_producers <- top_5_producers %>% 
  arrange(master_year)

#write_csv(top_5_producers, "oil_loss_by_company.csv")
```

### The companies who have lost the most since 2009

Amid Silver Dollar was incorporated in 2013, yet they did not log any losses until 2019. 

```{r}
top_5_losses <- oil_loss %>% 
  filter(operator_number=="NFR ENERGY, LLC"|operator_number=="AMID SILVER DOLLAR PIPELINE LLC"|operator_number=="ENDEAVOR ENERGY RESOURCES L.P."|operator_number=="EPIC CONSOLIDATED OPS, LLC"|operator_number=="ANADARKO E&P ONSHORE LLC") %>%
  group_by(operator_number, master_year) %>% 
  summarize(total_lost = sum(net_loss, na.rm=TRUE))

top_5_losses <- top_5_losses %>% 
  pivot_wider(names_from=operator_number, values_from=total_lost)
 
top_5_losses <- top_5_losses %>% 
  arrange(master_year)
```

### Question: why does Pioneer Natural Resources have so many incidents?

Interesting how there have been 567 recorded instances of oil loss due to equipment failure, yet only 5 of the 769 incidents had "installed new equipment" or "repaired existing equipment" in the remedy column.

```{r}
oil_loss %>% 
  filter(operator_number=="PIONEER NATURAL RES. USA, INC.") %>% 
  group_by(remedy) %>% 
  summarize(total=n())

oil_loss %>%
  filter(operator_number=="PIONEER NATURAL RES. USA, INC.") %>% 
  group_by(causeof_loss) %>% 
  summarize(total=n())
```

## Section 3: The Affected Counties

### breakdown by county

Oil loss has been reported in 204 Texas counties. View of all fracking locations in Texas: <https://ft.maps.arcgis.com/apps/webappviewer/index.html?appid=f537aa46be49431fbb820c4e99e63474> Annual oil production by county: <https://www.rrc.texas.gov/resource-center/data-visualization/oil-gas-data-visualization/oil-and-gas-production/>

<iframe title="Crockett County and Harrison County most affected by Texas oil company barrel losses" aria-label="Map" id="datawrapper-chart-ogAF0" src="https://datawrapper.dwcdn.net/ogAF0/2/" scrolling="no" frameborder="0" style="width: 0; min-width: 100% !important; border: none;" height="826" data-external="1">

</iframe>

```{=html}
<script type="text/javascript">!function(){"use strict";window.addEventListener("message",(function(a){if(void 0!==a.data["datawrapper-height"]){var e=document.querySelectorAll("iframe");for(var t in a.data["datawrapper-height"])for(var r=0;r<e.length;r++)if(e[r].contentWindow===a.source){var i=a.data["datawrapper-height"][t]+"px";e[r].style.height=i}}}))}();
</script>
```

```{r}
counties <- oil_loss %>% 
  group_by(county) %>% 
  summarize(total_lost = sum(net_loss, na.rm=TRUE)) %>% 
  arrange(desc(total_lost))

#write_csv(counties, "texas_counties.csv")

oil_loss %>% 
  group_by(county) %>% 
  summarize(total_incidents = n()) %>%
  arrange(desc(total_incidents))

#midland county
oil_loss %>% 
  filter(county=="MIDLAND") %>% 
  group_by(master_year) %>% 
  summarize(total_lost = sum(net_loss, na.rm=TRUE))

#crockett county
oil_loss %>% 
  filter(county=="CROCKETT") %>% 
  group_by(master_year) %>% 
  summarize(total_lost = sum(net_loss, na.rm=TRUE))

#crockett has the highest total lost, midland has the most incidents
```

## Section 4: Causes for the Losses

```{r}
oil_loss %>% 
  group_by(causeof_loss) %>% 
  summarize(total=n()) %>% 
  arrange(desc(total))

equip_failures <- oil_loss %>% 
  select(master_year, causeof_loss) %>% 
  filter(causeof_loss=="Equipment Failure"|causeof_loss=="Corrosion - Rust") %>%
  group_by(master_year) %>% 
  summarize(total = n())
  
oil_loss %>% 
  filter(causeof_loss=="Equipment Failure"|causeof_loss=="Corrosion - Rust") %>% 
  filter(remedy=="Repaired Existing Equipment") %>% 
  count()
```
